Enforcing Idempotency in Neural Networks

### 2025-01-31 paper.tex
For submission

% Use the following line for the initial blind version submitted for review:
%\usepackage{icml2025}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2025}

@@

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Enforcing idempotency in neural networks}

@@

\twocolumn[
    \icmltitle{Enforcing idempotency in neural networks}

@@

% Wider context -> Idempotent networks
Using neural networks as data augmentation tools is becoming more widespread in areas such as signal processing and generative artificial intelligence. In particular, networks of the form ${f: X \to X}$, mapping data within the same space $X$, are frequently used in image augmentation \cite{lu-image-aug}, video generation \cite{ma-compression, liu-gans}, sorting algorithms \cite{tambouratzis-sorting}, compression algorithms \cite{namphol-compression, liu-gans}, image denoising \cite{mao-deblurring, ilesanmi-denoising, liu-gans}, and image generation \cite{liu-gans}, among others.

Some data transformation tasks admit \textit{only} idempotent solutions (\textit{e.g.}, sorting), whilst other tasks admit \textit{no} idempotent solutions (\textit{e.g.}, rotating an image by 90$^{\circ}$). This work is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in Section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (\textit{e.g.}, application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{mao-deblurring,liu-gans}. Since solutions are not intrinsically idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

@@

Although we assume $\vK$ is close to idempotent, we show that in practice $g$ can be used to derive matrices which are within machine precision of perfect idempotency even when the input matrix $\vK$ is relatively far from idempotent. At a high level, this process is based on a recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$, taking small $\gamma$-sized steps in the direction of $g(\vK)$. While this recurrence relation derives idempotent matrices -- and can therefore be used to train single-layer networks with identity activations to be idempotent -- we also give a more general application of Eq. (\ref{eq:g}) as a modification of the backpropagation algorithm, yielding an architecture agnostic and efficient algorithm for finding idempotent networks. As we will see, this modification in general not only leads to higher idempotent error reduction but also seems to explore the loss landscape differently from the canonical approach.

% Outline of each section of the paper
In Section \ref{sec:method-idea} we give a detailed description of the method used to derive Eq. (\ref{eq:g}) and alternative solutions. We also show that while there exists non-idempotent fixed points to Eq. (\ref{eq:g}), these points are repelling under the recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$ for $0 \leq \gamma \leq 1$, giving credence to the use of such a recurrence relation in practice. Finally, in Section \ref{sec:method-scheme} we derive a full training scheme for training arbitrary neural network architectures of the form ${f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n}$. In Section \ref{sec:experiment}, we present experimental data for a variety of fully-connected network architectures, showing that our method outperforms ordinary backpropagation under varied conditions. We also replicate the results of \citealt{shocher-ign} by applying our method on a U-net style DCGAN network to successfully create a generative network for the MNIST dataset. Lastly, Sections \ref{sec:related} and \ref{sec:conclusion} discuss how our method distinguishes itself from related approaches as well as future work.

@@

In the case when $n=1$ we consider ${\vD^2 \approx \bm{0}}$ and the expression ${(\vK')^2 - \vK' = \bm{0}}$ can be expanded and reduced by recursively applying the following assumptions, for all $\vX,\vY,\vZ$ matrices:
%
\begin{align}
    \vD^2 \approx \bm{0}, \quad \vP^2 = \vP, \quad \vX \vD \vY \vD \vZ \approx \bm{0}.
\end{align}
%
When $j\leq2$, \textit{i.e.}, there exists no solutions for $\alpha_i$. When $j=3$, however, there is exactly one solution when ${\alpha_1 = 0}$, ${\alpha_2 = 3}$ and ${\alpha_3 = -2}$, thus $g$ is as defined in Eq. (\ref{eq:g}). For $j>3$ there exists families of solutions (see Appendix \ref{app:solutions}), but we consider primarily the case when $j=3$ as this requires fewer higher-order terms of $\vK$ and is therefore generally less costly to compute.

@@

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
        \begin{axis}[
                axis lines = middle,
                xlabel = {$\vK$},
                ylabel = {$\vK'$},
                domain=-0.75:1.75,
                samples=100,
                grid=both,
            ]
            \addplot[
                color=blue,
                thick
            ]{3*x^2-2*x^3};
        \end{axis}
    \end{tikzpicture}
    \caption{Plot of $\vK' = 3 \vK^2 - 2 \vK^3$ in the case $\vK$ is scalar.}
    \label{fig:plot-g}
\end{figure}

In general, we place no restrictions on the matrix ${\vK \in \mathbb{R}^{m \times m}}$. In particular, it might not be directly diagonalizable. It is well known, however, that for every square matrix $\vK$ there exists an invertible matrix $\vP$ and a Jordan normal form \cite{jordan-form} $\vJ \in \mathbb{C}^{m \times m}$ of ${\vK \in \mathbb{R}^{m \times m}}$ such that ${\vK = \vP \vJ \vP^{-1}}$. From this the dual problem,
\begin{align}
    \vJ = 3 \vJ^2 - 2 \vJ^3,
\end{align}
can be constructed. Let $d_\lambda$ denotes the algebraic multiplicity of eigenvalue $\lambda$ of $\vK$. Then, the block-diagonal structure of $\vJ$ imposes up to four equations per eigenvalue of $\vK$ (see Appendix \ref{app:jordan}):
\begin{align}
    \lambda & = 3\lambda^2 - 2\lambda^3 & \label{eq:cond1}                                   \\
    1       & = 6\lambda - 6\lambda^2   & \text{Only when $d_\lambda\geq2$.}\label{eq:cond2} \\
    0       & = 3 - 6\lambda            & \text{Only when $d_\lambda\geq3$.}\label{eq:cond3} \\
    0       & = 0 - 2                   & \text{Only when $d_\lambda\geq4$.}\label{eq:cond4}
\end{align}

@@

Whilst this only technically applies in the linearized setting, we propose to also apply the method in non-linear settings using the following recurrence relation, for $0 \leq \gamma \leq 1$:
%
\begin{align}
    \vK' = \vK + \gamma(g(\vK) - \vK).
\end{align}
%
This has the effect of taking small $\gamma$-sized steps in the direction of $g$ at every time point.

@@

Consider a neural network ${f_{\vtheta}: \mathbb{R}^m \to \mathbb{R}^m}$ together with its application to input ${\vx \in \mathbb{R}^m}$, denoted ${\vy = f_{\vtheta}(\vx)}$. We might then consider the recurrence relation in Eq. (\ref{eq:g}) in the following form:
%
\begin{align}
    \vy' = 3f_{\vtheta}(\vy) - 2f_{\vtheta}(f_{\vtheta}(\vy))
\end{align}
%
This describes a desired change in the output of the network which we denote ${\Delta f_{\vtheta}(\vx) = \vy' - \vy}$. In other words, ${\Delta f_{\vtheta}(\vx)}$ describes the change in $\vy$ which moves $\vy$ towards an idempotent projection much in the same way that the quantity $\pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy}$ describes the direction which reduces the idempotent loss function $\mathcal{L}_{\mathrm{idem}}(\vy)$ in Eq. (\ref{eq:idem-loss}). A central idea presented in this work is therefore the definition
%
\begin{align}
    \pd{(-\mathcal{L}(\vy))}{\vy} \equiv \Delta f_{\vtheta}(\vx)
    \label{eq:substitution}
\end{align}
%
as an alternative quantity to the traditional, analytical solution to $\pd{(-\mathcal{L}(\vy))}{\vy}$. To complete the scheme, we consider how a change in the output $\vy$ can be propagated to a change in the parameters $\vtheta$ of $f_{\vtheta}$. This, however, is a straightforward application of the chain rule as it is calculated conventionally in backpropagation. We use the term ``\textbf{Modified Backpropagation}'' to refer to the canonical backpropagation algorithm with the rule (\ref{eq:substitution}) applied appropriately when computing gradients.

@@

% (3 pages)
\section{Experimental Results}
\label{sec:experiment}
%\textit{On experimental networks we give line graphs comparing absolute error at varying learning rates. We also give line graphs for LRs over many epochs. Takeaway message is that for deeper/wider networks we outperform ordinary autodiff by a lot.}
% Outline differences in training scheme.
% Error across LR for B3&B4
% Error across Runs for B3&B4
% (PCA analysis graphs)
To evaluate the training scheme suggested in Section \ref{sec:method-scheme} we compare relative performance between the two methods: ``Ordinary Backpropagation'' with the quantity $\pd{(-\mathcal{L}(\vy))}{\vy}$ resolved at runtime by automatic differentiation, and ``Modified Backpropagation'' with the modified backpropagation rule for $\pd{(-\mathcal{L}(\vy))}{\vy}$. To demonstrate the flexibility of the approach, we report results for four diverse MLP-style networks, as described in table \ref{tab:networks}.

@@

\begin{table}[htbp]
    \begin{center}
        \begin{tabular}{|c|p{0.18\textwidth}|c|}
            \hline
            \textbf{Identifier} & \textbf{Architecture}                                                                                                    & \textbf{No. Parameters} \\
            \hline
            \hline
            B1                  & Linear(5, 5)                                                                                                             & $30$                    \\
            \hline
            B2                  & Linear(128, 256) \newline Linear(256, 256) \newline Linear(256, 256) \newline Linear(256, 256) \newline Linear(256, 128) & $263\,296$              \\
            \hline
            B3                  & Linear(4096, 1024) \newline Linear(1024, 4096)                                                                           & $8\,393\,728$           \\
            \hline
            B4                  & Linear(784, 1024) \newline Linear(1024, 2048) \newline Linear(2048, 784)                                                 & $4\,509\,456$           \\
            \hline
        \end{tabular}
        \vspace{5pt}
        \caption{Four neural networks for testing. Each ``Linear($n$, $m$)'' block is parameterized by its input dimension $n$ and its output dimension $m$, corresponding to the underlying $\vW \in \mathbb{R}^{m \times n}$ weight matrix. Every block has an associated bias vector and LeakyReLU($0.2$) activation function. B1 represents a trivial network, B2 represents a relatively deep network, B3 represents a relatively wide network, and B4 represents a more realistic network.}
        \label{tab:networks}
    \end{center}
\end{table}

@@

\subsection{Qualitative differences}
\label{sec:experiment-qual}
%\textit{TODO: Include a section on qualitative differences in behaviour of modified and ordinary backpropagation. Does one method travel a different route in the loss landscape? This small section is to be suggestive of distinguishing features of the new method.}
%
In this section we present suggestive evidence that Modified Backpropagation searches the solution space differently from Ordinary Backpropagation.
%
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\columnwidth]{./resources/pca_b2.png}
    \caption{Representative projections of the optimizer trajectories over 2500 epochs of either algorithm on the B2 model at optimal learning rates (Figure \ref{fig:avg-abs-err}). Total variance captured is $>97.8\%$ with cosine similarity of PC1 and PC2 less than $1.0\times10^{-6}$. Optimizer trajectory of Modified Backpropagation deviates significantly from Ordinary Backpropagation.}
    \label{fig:pca-b2}
\end{figure}

@@

Qualitative evaluation show that Modified and Ordinary Backpropagation often differ significantly in projected trajectories across the two most explanatory directions, but this is not always the case (\textit{e.g.}, B4 in Figure \ref{fig:big-pca}). Additionally, optimization trajectories for Modified Backpropagation can be explained by projection onto two direction with more than $90\%$ variance explained, indicating that it exhibits the same behaviour as Ordinary Backpropagation which has previously been suggested to largely operate in low-dimensional subspaces \cite{li-visualizing,song-subspaces}. One should note, however, that the loss surface is here represented under a dramatic dimensionality reduction which makes further conclusions unreliable.

@@

We now investigate how the gradients produced by Modified Backpropagation differ from those produced by Ordinary Backpropagation. We give here an analysis over a single training run on network B2, but similar results hold for all networks in table \ref{tab:networks} over repetitions of the experiment. As Figure \ref{fig:cos-b2} shows, gradients suggested by either algorithm remain relatively dissimilar throughout training, which further indicates a difference in the expected optimization trajectory. Furthermore, as evidenced by Figures \ref{fig:pca-b2} and \ref{fig:cos-b2}, Modified Backpropagation travels faster (\textit{i.e.}, gives stronger gradient signal) than Ordinary Backpropagation, even when optimal learning rates are selected for both algorithms.

@@

\subsection{Quantitative differences}
\label{sec:experiment-quant}
We now give an evaluation of the relative efficacy of Modified Backpropagation to Ordinary Backpropagation. As shown in Figure \ref{fig:avg-abs-err}, for networks B2-B4 Modified Backpropagation achieves significantly lower absolute idempotent error on average at lower learning rate. For network B3 the difference is more than one order of magnitude. As the tested networks represent varying architectures with a commonly used activation function, these results suggest that Modified Backpropagation fares well in a variety of training configurations.

@@

% There is future work in uncovering the relationship between the suggested gradient from MB and that of OB
\subsection{Application to Generative Networks}
\label{sec:experiment-gen}
%\textit{We replicate the results of IGN on MNIST and CelebA. Latent space analysis. Takeaway is that we show application to a U-net GAN architecture based on Conv layers.}
As mentioned, one of the motivating factors for actively enforcing idempotency during training is to apply it as a secondary optimization objective in conjunction with optimizing for a primary task. In this section we replicate the results of \citealt{shocher-ign} as we train a U-net style DCGAN architecture (see Appendix \ref{app:gen-training-scheme}) on the MNIST dataset. Let $\mathcal{D}$ denote the distribution of MNIST samples, while ${\mathcal{D}' = \mathcal{N}(0, 1)}$ from which noise is sampled. Let $\vtheta'$ is a copy of the trainable weights $\vtheta$ at each time step, where $\vtheta'$ is detached from the computational graph. In this training scheme, the loss function being optimized is
%
\begin{align}
    \begin{split}
        \mathcal{L}{(\vtheta, \vtheta')} & = \lambda_r \mathcal{L}_{\mathrm{rec}}{(\vtheta)}                                                                           \\
                                         & + \lambda_i \mathcal{L}_{\mathrm{idem}}{(\vtheta, \vtheta')} + \lambda_r \mathcal{L}_{\mathrm{tight}}{(\vtheta, \vtheta')}.
    \end{split}
    \label{eq:gan-loss}
\end{align}
%
To see why employing two copies of the weights is useful, consider $(\vx, \vy^*) \sim \mathcal{D}$ and $\vz \sim \mathcal{D}'$ and the individual loss components:
%
\begin{align}
    \mathcal{L}_{\mathrm{rec}}{((\vx, \vy^*); \vtheta)}    & = \Vert \vy^* - f_{\vtheta}(\vx) \Vert_1                            \\
    \mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}  & = \Vert f_{\vtheta'}(f_{\vtheta}(\vz)) - f_{\vtheta}(\vz) \Vert_1   \\
    \mathcal{L}_{\mathrm{tight}}{(\vz; \vtheta, \vtheta')} & = -\Vert f_{\vtheta}(f_{\vtheta'}(\vz)) - f_{\vtheta'}(\vz) \Vert_1
\end{align}
%
For instance, the quantity $\pd{\mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}}{\vtheta}$ is only affected by the inner application of $f$ above due to $\vtheta'$ being detached from the graph. The relationship between loss components $\mathcal{L}_{\mathrm{idem}}$ and $\mathcal{L}_{\mathrm{tight}}$ is adversarial in nature.

The major difference in this work from \citealt{shocher-ign} is that we use Modified Backpropagation for implementing both $\mathcal{L}_{\mathrm{idem}}$ and $\mathcal{L}_{\mathrm{tight}}$. As such, we effectively evaluate both loss components as the Mean Squared Error (MSE) instead of the $L_1$ loss, and we use the training scheme in Section \ref{sec:method-scheme} to construct the backwards pass (see Appendix \ref{app:gen-training-scheme}). We use the same implementation for $\mathcal{L}_{\mathrm{rec}}$ as above.

We have successfully replicated several results of \citealt{shocher-ign} under this training scheme. In particular, Figure \ref{fig:gen-mnist} shows qualitative examples of noise drawn from $\mathcal{D}'$ being mapped to images resembling samples from the MNIST dataset. While samples remains largely similar between the first and second application of the network, in some cases we do also observe the same ``self-correction'' of artefacts/noise after the second application observed by \citealt{shocher-ign}. In Figure \ref{fig:latent-mnist} we visualize the effect of applying the trained network to noise linearly interpolated between two clear samples $\vA, \vB \in \mathbb{R}^{28 \times 28}$. We again observe the secondary application of the network ``cleaning up'' images. For more uncurated examples of generated images, see Appendix \ref{app:gen-samples}.

@@

\subsection{Alternatives to gradients}
There is a large body of work focusing on approximating gradients. In \citealt{spall-perturb} a stochastic variant of finite differences called SPSA is suggested with competitive performance at lower computational cost. \citealt{scheinberg-approx,do-approx,bandler-approx} all further explore this and similar approaches based on finite differences. Nevertheless, a core idea of Modified Backpropagation is the direct substitution of the gradient of $\mathcal{L}_{\mathrm{idem}}$ with the quantity in Eq. (\ref{eq:substitution}). Although the exact relationship between this quantity and the gradients produced by Ordinary Backpropagation is still unclear, it is certainly not a direct approximation as evidenced in Section \ref{sec:experiment-qual}.

>>>>

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2025}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2025}

@@

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Enforcing Idempotency in Neural Networks}

@@

\twocolumn[
    \icmltitle{Enforcing Idempotency in Neural Networks}

@@

% Wider context -> Idempotent networks
Using neural networks as data augmentation tools is becoming more widespread in areas such as signal processing and generative artificial intelligence. In particular, networks of the form ${f: X \to X}$, mapping data within the same space $X$, are frequently used in image augmentation \cite{lu-image-aug}, video generation \cite{ma-compression, liu-gans}, sorting algorithms \cite{tambouratzis-sorting}, compression algorithms \cite{namphol-compression, liu-gans}, image denoising \cite{ilesanmi-denoising, liu-gans, mao-deblurring}, and image generation \cite{liu-gans}, among others.

Some data transformation tasks admit \textit{only} idempotent solutions (\textit{e.g.}, sorting), whilst other tasks admit \textit{no} idempotent solutions (\textit{e.g.}, rotating an image by 90$^{\circ}$). This work is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in Section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (\textit{e.g.}, application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{liu-gans,mao-deblurring}. Since solutions are not inherently idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

@@

Although we assume $\vK$ is close to idempotent, we show that in practice $g$ can be used to derive matrices which are within machine precision of perfect idempotency even when the input matrix $\vK$ is relatively far from idempotent. At a high level, this process is based on a recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$, taking small $\gamma$-sized steps in the direction of $g(\vK)$. While this recurrence relation derives idempotent matrices -- and can therefore be used to train single-layer networks with identity activations to be idempotent -- we also give a more general application of Eq. (\ref{eq:g}) as a modification of the backpropagation algorithm, yielding an architecture agnostic and efficient algorithm for finding idempotent networks. As we will see, this modification in general not only leads to significantly improved idempotent error reduction but also explores the loss landscape differently from the canonical approach.

% Outline of each section of the paper
In Section \ref{sec:method-idea} we give a detailed description of the method used to derive Eq. (\ref{eq:g}) and alternative solutions. We also show that while there exists non-idempotent fixed points to Eq. (\ref{eq:g}), these points are repelling under the recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$ for $0 \leq \gamma \leq 1$, giving credence to the use of such a recurrence relation in practice. Finally, in Section \ref{sec:method-scheme} we derive a full training scheme for training arbitrary neural network architectures of the form ${f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n}$. In Section \ref{sec:experiment}, we present experimental data for a variety of fully-connected network architectures, showing that our method outperforms ordinary backpropagation under varied conditions. We also replicate the results of \citealt{shocher-ign} by applying our method on a U-net style DCGAN model to successfully create a generative network for the MNIST dataset. Lastly, Sections \ref{sec:related} and \ref{sec:conclusion} discuss how our method distinguishes itself from related approaches as well as future work.

@@

In the case when $n=1$ we consider ${\vD^2 \approx \bm{0}}$ and the expression ${(\vK')^2 - \vK' = \bm{0}}$ can be expanded and reduced by recursively applying the following assumptions, for all $\vX,\vY,\vZ$ matrices:
%
\begin{align}
    \vD^2 \approx \bm{0}, \quad \vP^2 = \vP, \quad \vX \vD \vY \vD \vZ \approx \bm{0}.
\end{align}
%
When $j\leq2$, there exists no solutions for $\alpha_i$. When $j=3$ there is exactly one solution when ${\alpha_1 = 0}$, ${\alpha_2 = 3}$ and ${\alpha_3 = -2}$, which gives precisely $g$ as defined in Eq. (\ref{eq:g}). For $j>3$ there exists families of solutions (see Appendix \ref{app:solutions}), but we consider primarily the case when $j=3$ as this requires fewer higher-order terms of $\vK$ and is therefore generally less costly to compute.

@@

\begin{figure}[t]
    \centering
    \begin{tikzpicture}
        \begin{axis}[
                axis lines = middle,
                xlabel = {$\vK$},
                ylabel = {$\vK'$},
                domain=-0.75:1.75,
                samples=100,
                grid=both,
            ]
            \addplot[
                color=blue,
                thick
            ]{3*x^2-2*x^3};
        \end{axis}
    \end{tikzpicture}
    \caption{Plot of $\vK' = 3 \vK^2 - 2 \vK^3$ in the case $\vK$ is scalar.}
    \label{fig:plot-g}
\end{figure}

In general, we place no restrictions on the matrix ${\vK \in \mathbb{R}^{m \times m}}$. In particular, it might not be directly diagonalizable. It is well known, however, that for every square matrix $\vK$ there exists an invertible matrix $\vP$ and a Jordan normal form \cite{jordan-form} $\vJ \in \mathbb{C}^{m \times m}$ of ${\vK \in \mathbb{R}^{m \times m}}$ such that ${\vK = \vP \vJ \vP^{-1}}$. From this the dual problem,
\begin{align}
    \vJ = 3 \vJ^2 - 2 \vJ^3,
\end{align}
can be constructed. Let $d_\lambda$ denote the algebraic multiplicity of eigenvalue $\lambda$ of $\vK$. Then, the block-diagonal structure of $\vJ$ imposes up to four equations per eigenvalue of $\vK$ (see Appendix \ref{app:jordan}):
\begin{align}
    \lambda & = 3\lambda^2 - 2\lambda^3 & \label{eq:cond1}                                   \\
    1       & = 6\lambda - 6\lambda^2   & \text{Only when $d_\lambda\geq2$.}\label{eq:cond2} \\
    0       & = 3 - 6\lambda            & \text{Only when $d_\lambda\geq3$.}\label{eq:cond3} \\
    0       & = 0 - 2                   & \text{Only when $d_\lambda\geq4$.}\label{eq:cond4}
\end{align}

@@

Whilst this analysis technically only applies in the linearized setting, we propose to also apply the method in non-linear settings using the following recurrence relation, for ${0 \leq \gamma \leq 1}$:
%
\begin{align}
    \vK' = \vK + \gamma(g(\vK) - \vK).
\end{align}
%
This has the effect of taking small $\gamma$-sized steps in the direction of $g$ at every time point.

@@

Consider a neural network ${f_{\vtheta}: \mathbb{R}^m \to \mathbb{R}^m}$ together with its application to input ${\vx \in \mathbb{R}^m}$, denoted ${\vy = f_{\vtheta}(\vx)}$. We might then consider the recurrence relation in Eq. (\ref{eq:g}) in the following form:
%
\begin{align}
    \vy' = 3f_{\vtheta}(\vy) - 2f_{\vtheta}(f_{\vtheta}(\vy))
\end{align}
%
This describes a desired change in the output of the network which we denote ${\Delta f_{\vtheta}(\vx) = \vy' - \vy}$. In other words, ${\Delta f_{\vtheta}(\vx)}$ describes the change in $\vy$ which moves $\vy$ towards an idempotent projection much in the same way that the quantity $\pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy}$ describes the direction which reduces the idempotent loss function $\mathcal{L}_{\mathrm{idem}}(\vy)$ in Eq. (\ref{eq:idem-loss}). A central idea presented in this work is therefore the definition
%
\begin{align}
    \pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy} \equiv \Delta f_{\vtheta}(\vx)
    \label{eq:substitution}
\end{align}
%
as an alternative quantity to the traditional, analytical solution to $\pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy}$. To complete the scheme, we consider how a change in the output $\vy$ can be propagated to a change in the parameters $\vtheta$ of $f_{\vtheta}$. This, however, is a straightforward application of the chain rule as it is calculated conventionally in backpropagation. We use the term ``\textbf{Modified Backpropagation}'' to refer to the canonical backpropagation algorithm with the rule (\ref{eq:substitution}) applied appropriately when computing gradients.

@@

% (3 pages)
\section{Experimental Results}
\label{sec:experiment}
%\textit{On experimental networks we give line graphs comparing absolute error at varying learning rates. We also give line graphs for LRs over many epochs. Takeaway message is that for deeper/wider networks we outperform ordinary autodiff by a lot.}
% Outline differences in training scheme.
% Error across LR for B3&B4
% Error across Runs for B3&B4
% (PCA analysis graphs)
To evaluate the training scheme suggested in Section \ref{sec:method-scheme} we compare relative performance between the two methods: ``Ordinary Backpropagation'' with the quantity $\pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy}$ resolved at runtime by automatic differentiation, and ``Modified Backpropagation'' with the modified backpropagation rule for $\pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy}$. To demonstrate the flexibility of the approach, we report results for four diverse MLP-style networks, as described in Table \ref{tab:networks}.

@@

\begin{table}[htbp]
    \begin{center}
        \caption{Four neural networks for testing. Each ``Linear($n$, $m$)'' block is parameterized by its input dimension $n$ and its output dimension $m$, corresponding to the underlying $\vW \in \mathbb{R}^{m \times n}$ weight matrix. Every block has an associated bias vector and LeakyReLU($0.2$) activation function. B1 represents a trivial network, B2 represents a relatively deep network, B3 represents a relatively wide network, and B4 represents a more realistic network.}
        \label{tab:networks}
        \vspace{0.1in}
        \begin{tabular}{|c|p{0.18\textwidth}|c|}
            \hline
            \textbf{Identifier} & \textbf{Architecture}                                                                                                    & \textbf{No. Parameters} \\
            \hline
            \hline
            B1                  & Linear(5, 5)                                                                                                             & $30$                    \\
            \hline
            B2                  & Linear(128, 256) \newline Linear(256, 256) \newline Linear(256, 256) \newline Linear(256, 256) \newline Linear(256, 128) & $263\,296$              \\
            \hline
            B3                  & Linear(4096, 1024) \newline Linear(1024, 4096)                                                                           & $8\,393\,728$           \\
            \hline
            B4                  & Linear(784, 1024) \newline Linear(1024, 2048) \newline Linear(2048, 784)                                                 & $4\,509\,456$           \\
            \hline
        \end{tabular}
    \end{center}
\end{table}

@@

\subsection{Qualitative Differences}
\label{sec:experiment-qual}
%\textit{TODO: Include a section on qualitative differences in behaviour of modified and ordinary backpropagation. Does one method travel a different route in the loss landscape? This small section is to be suggestive of distinguishing features of the new method.}
%
In this section we present suggestive evidence that Modified Backpropagation searches the solution space differently from Ordinary Backpropagation.
%
\begin{figure}[t]
    \centering
    \includegraphics[width=0.85\columnwidth]{./resources/pca_b2.png}
    \caption{Representative projections of the optimizer trajectories over 2500 epochs of either algorithm on the B2 model at optimal learning rates (Figure \ref{fig:avg-abs-err}). Total variance captured is $>97.8\%$ with cosine similarity of PC1 and PC2 less than $1.0\times10^{-6}$. Optimizer trajectory of Modified Backpropagation deviates significantly from Ordinary Backpropagation.}
    \label{fig:pca-b2}
\end{figure}

@@

Qualitative evaluation show that Modified and Ordinary Backpropagation often differ significantly in projected trajectories across the two most explanatory directions, but this is not always the case (\textit{e.g.}, B4 in Figure \ref{fig:big-pca}). Additionally, optimization trajectories for Modified Backpropagation can be explained by projection onto two direction with more than $90\%$ variance explained, indicating that it exhibits the same behaviour as Ordinary Backpropagation which has previously been suggested to largely operate in low-dimensional subspaces \cite{li-visualizing,song-subspaces}. One should note, however, that the loss surface is here represented under a dramatic dimensionality reduction which limits further conclusions.

@@

We now investigate how the gradients produced by Modified Backpropagation differ from those produced by Ordinary Backpropagation. We give here an analysis over a single training run on network B2, but similar results hold for all networks in Table \ref{tab:networks} over repetitions of the experiment. As Figure \ref{fig:cos-b2} shows, gradients suggested by either algorithm remain relatively dissimilar throughout training, which further indicates a difference in the expected optimization trajectory. Furthermore, as evidenced by Figures \ref{fig:pca-b2} and \ref{fig:norm-b2}, Modified Backpropagation travels faster (\textit{i.e.}, gives stronger gradient signal) than Ordinary Backpropagation, even when optimal learning rates are selected for both algorithms.

@@

\subsection{Quantitative Differences}
\label{sec:experiment-quant}
We now give an evaluation of the relative efficacy of Modified Backpropagation to Ordinary Backpropagation. As shown in Figure \ref{fig:avg-abs-err}, for networks B2-B4 Modified Backpropagation achieves significantly lower absolute idempotent error on average at lower learning rate. For network B3 the difference is more than one order of magnitude. As the tested networks represent varying architectures with a commonly used activation function, these results suggest that Modified Backpropagation fares well in a variety of training configurations.

@@

% There is future work in uncovering the relationship between the suggested gradient from MB and that of OB
\subsection{Application to Generative Networks}
\label{sec:experiment-gen}
%\textit{We replicate the results of IGN on MNIST and CelebA. Latent space analysis. Takeaway is that we show application to a U-net GAN architecture based on Conv layers.}
As mentioned, one of the motivating factors for actively enforcing idempotency during training is to apply it as a secondary optimization objective in conjunction with optimizing for a primary task. In this section we replicate the results of \citealt{shocher-ign} as we train a U-net style DCGAN architecture (see Appendix \ref{app:gen-training-scheme}) on the MNIST dataset. Let $\mathcal{D}$ denote the distribution of MNIST samples, while ${\mathcal{D}' = \mathcal{N}(0, 1)}$ from which noise is sampled. Let $\vtheta'$ be a copy of the trainable weights $\vtheta$ at each time step, where $\vtheta'$ is detached from the computational graph. In this training scheme, the loss function being optimized is
%
\begin{align}
    \begin{split}
        \mathcal{L}{(\vtheta, \vtheta')} & = \lambda_r \mathcal{L}_{\mathrm{rec}}{(\vtheta)}                                                                           \\
                                         & + \lambda_i \mathcal{L}_{\mathrm{idem}}{(\vtheta, \vtheta')} + \lambda_r \mathcal{L}_{\mathrm{tight}}{(\vtheta, \vtheta')}.
    \end{split}
    \label{eq:gan-loss}
\end{align}
%
To see why employing two copies of the weights is useful, consider $(\vx, \vy^*) \sim \mathcal{D}$ and $\vz \sim \mathcal{D}'$ and the individual loss components:
%
\begin{align}
    \mathcal{L}_{\mathrm{rec}}{((\vx, \vy^*); \vtheta)}    & = \Vert \vy^* - f_{\vtheta}(\vx) \Vert_1                            \\
    \mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}  & = \Vert f_{\vtheta'}(f_{\vtheta}(\vz)) - f_{\vtheta}(\vz) \Vert_1   \\
    \mathcal{L}_{\mathrm{tight}}{(\vz; \vtheta, \vtheta')} & = -\Vert f_{\vtheta}(f_{\vtheta'}(\vz)) - f_{\vtheta'}(\vz) \Vert_1
\end{align}
%
For instance, the quantity $\pd{\mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}}{\vtheta}$ is only affected by the inner application of $f$ above due to $\vtheta'$ being detached from the graph. The relationship between loss components $\mathcal{L}_{\mathrm{idem}}$ and $\mathcal{L}_{\mathrm{tight}}$ is adversarial in nature.

The major difference in this work from \citealt{shocher-ign} is that we use Modified Backpropagation for implementing both $\mathcal{L}_{\mathrm{idem}}$ and $\mathcal{L}_{\mathrm{tight}}$. As such, we \textit{evaluate} both loss components as the Mean Squared Error (MSE) instead of the $L_1$ loss, and we use the training scheme in Section \ref{sec:method-scheme} to perform the backwards pass (see also Appendix \ref{app:gen-training-scheme}). We use the same implementation for $\mathcal{L}_{\mathrm{rec}}$ as above.

We have successfully replicated several results of \citealt{shocher-ign} under this training scheme. In particular, Figure \ref{fig:gen-mnist} shows qualitative examples of noise drawn from $\mathcal{D}'$ being mapped to images resembling samples from the MNIST dataset. While samples remain largely similar between the first and second application of the network, in some cases we do also observe the same ``self-correction'' of artefacts/noise after the second application observed by \citealt{shocher-ign}. In Figure \ref{fig:latent-mnist} we visualize the effect of applying the trained network to noise linearly interpolated between two clear samples $\vA, \vB \in \mathbb{R}^{28 \times 28}$. We again observe the secondary application of the network ``cleaning up'' images. For more uncurated examples of generated images, see Appendix \ref{app:gen-samples}.

@@

\subsection{Alternatives to Gradients}
There is a large body of work focusing on approximating gradients. In \citealt{spall-perturb}, a stochastic variant of finite differences called SPSA is suggested with competitive performance at lower relative computational cost. \citealt{bandler-approx,do-approx,scheinberg-approx} all further explore this and similar approaches based on finite differences. Nevertheless, a core idea of Modified Backpropagation is the direct substitution of the gradient of $\mathcal{L}_{\mathrm{idem}}$ with the quantity in Eq. (\ref{eq:substitution}). Although the exact relationship between this quantity and the gradients produced by Ordinary Backpropagation is still unclear, it is certainly not a trivial approximation as evidenced in Section \ref{sec:experiment-qual}.
